\vspace{-2.0em} Move Schedules: Fast persistence computations in coarse dynamic settings \thanks{ \textbf{Funding:} This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758. }

### 2023-01-24 move_schedules.tex
Major edits/simplifications to the 2-parameter section. Moving things to the appendix to shorten the paper

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices at thousands of snapshots in time.
Acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner, Morozov, and Edelsbrunner refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{matrix reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time.
The main step in the reduction algorithm is a specific decomposition $R = D V$ of the boundary matrix $D$ of $K$. 
Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
The vineyards algorithm transforms a time-varying filtration into an ordered set of permutations to be applied to adjacent rows and columns of the decomposition $D = RU$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for  homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is also efficient asymptotically: if there are $d$ time-points where the filtration order changes, then   vineyards   takes $O(m^3 + md)$ time; one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams independently at every time point.

@@

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all $d$ persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

@@

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total number of column operations) of various approaches.
  Since it is unknown a priori at which times  the persistent homology will change, one solution is to discretize the time domain at $n$ evenly spaced points and compute the persistence diagrams at each one of them  independently. 
  An alternative approach is to construct a (e.g., straight-line) homotopy between  contiguous time-points, and then decompose these homotopies into adjacent transpositions in the filtration order---the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. 
  The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. 
  As shown in the figure, if one needs to simulate the homotopy exactly, then the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

@@

\subsection{Main results}
Given a filtered simplicial complex $K$, denote by $R = D V$ the decomposition of its corresponding boundary matrix $D$ such that $R$ is reduced and $V$ is upper-triangular (see section~\ref{sec:reduction} for details).
 If one has a pair of filtrations $K, K'$ of equal size $m = \lvert K \rvert = \lvert K' \rvert$ and $R = D V$ has been computed, then it may be advantageous to use the information stored in ($R$, $V$) to reduce the computation of $R' = D' V'$. 
Given the appropriate permutation $P$ such that $D = P^T D P$, such an update scheme has the form: 
$$ (\ast P^T \ast R \ast P \ast) = (P^T DP)(\ast P^T \ast V \ast P \ast) $$
where $\ast$ is substituted with elementary column operations that repair the permuted decomposition. It is known how to generate $d \sim O(m^2)$ decompositions requiring on the order of $c \sim O(m^2)$~\cite{cohen2006vines} such column operations. Since each column operation take $O(m)$, the complexity of obtaining these $d$ decompositions takes $O(m^3)$. This is efficient if all $d$ decompositions are needed; otherwise, updating $R \mapsto R'$ matches the complexity of computing $R = D V$ independently.

We now summarize the main results obtained in this effort (Theorem~\ref{prop:sparsity_result}): suppose one has a sequence of permutations $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where each permutation $s_i$ is a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_\ast < j_\ast$, our first result is that this sequence can be computed in output-sensitive $O(d\kappa)$ time: 
\begin{equation}
	\quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert
\end{equation}
where the quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $K$ and $K'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself, which we take advantage of in Section~\ref{sec:schedule_cost}. Moreover, if the vineyards algorithm requires $C$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that our approach obtains the same sequence using at most $L$ column operations, where: 
\begin{equation}
	\frac{C}{2} \leq L \leq C
\end{equation}
Our second result turns towards the computation of $\mathcal{S}$ itself, and towards bounding its size $\lvert \mathcal{S} \rvert = d$. 
The complexity of determining $\mathcal{S}$ depends on the type of permutations considered; many edit distances exhibit tight quadratic growth rates and tend to require at least quadratic time (or worse) to compute~\cite{labarre2013lower}. 
We consider a special set of cyclic permutations whose sequences scale linearly in $m$ for all permutations in $S_m$: surprisingly, we find that the \emph{smallest} such sequence of these permutations which map $K \mapsto K'$ has size $d$ (Proposition~\ref{prop:schedule_size}), where: 
\begin{equation}
	d = m - \lvert \mathrm{LCS}(K, K') \rvert 
\end{equation}
Moreover, we also show that the information needed to construct any $\mathcal{S}$ with optimal size $\lvert \mathcal{S} \rvert = d$ can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. Although $d$ clearly is $O(m)$ for pathological inputs, we show that $d \sim m - \sqrt{m}$ in expectation for random filtrations (Corollary~\ref{cor:expectation}) is not too pessimistic for truly random filtrations, and we give empirical results suggesting $d$ is tends to be much smaller for time-varying filtrations.

In summary, we obtain an output-sensitive persistence update scheme which depends explicitly on the sparsity pattern of the involved matrices, has sub-cubic complexity in expectation even for random filtrations, and we show performs well in practice on a variety of applications.

@@

%\begin{proposition}\label{prop:factor2_main}
%	Let $R = D V$ be a valid decomposition of size $(m \times m)$, let $C$ denote the number of $O(m)$ operations required to execute schedule $S_F$, and similarly let $L$ be the number of $O(m)$ operations required to execute a coarsened schedule $\widetilde{S}_F$ of $S_F$. We have the following relationship between $C$ and $L$:
%	$$ \frac{C}{2} \leq L \leq C$$ 
%\end{proposition}

\subsection{Outline} The remainder of the paper is organized as follows: we review and establish the notations we will use to describe simplicial complexes, persistent homology, and dynamic persistence in Section \ref{sec:background}. 
We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 
In Section \ref{sec:results} we present applications of the proposed method. 
In particular, to the computation of crocker stacks from flock simulations, and to 2-dimensional persistence on the Mumford data set of image patches from natural images. 
In Section \ref{sec:conclusion} we conclude the paper by discussing other possible applications and future work.

@@

Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered   set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. There are two index sets of interest in what follows: $\mathbb{R}$ and $[n] = \{ 1, \dots, n\}$. 
Such a family is called a \emph{filtration},
which is deemed to be\emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ 
whenever $j$ is the immediate successor of $i$ in $I$.
Any finite filtration may be trivially converted into an simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
Without loss of generality, we exclusively consider simplexwise filtrations but for brevity-sake refer to them simply as filtrations.

@@

\noindent \textbf{Example 2.1:} Non-essential filtrations often arise in geometrical contexts. For example, given a finite metric space $(X, d)$, the \emph{Vietoris-Rips} complex at scale $\epsilon \in \mathbb{R}$ is the abstract simplicial complex given by: 

@@

A VR filtration indexed over $\mathbb{R}$  can be condensed to an essential filtration indexed over the set of pairwise distances $\{\, d(x,x') \; \mid \; x,x' \in X \, \}$, 
which can then be further reindexed to an essential simplexwise filtration by extending a total order of $X$ to its power set---e.g., 
using the lexicographical ordering. 

@@

Let $\mathbb{X}$ be a triangulable topological space. 
That is, so that there exists an abstract 
simplicial complex $K$ whose geometric realization  is homeomorphic to 
$\mathbb{X}$. 
Let  $f: \mathbb{X} \to \mathbb{R}$ be continuous and  
write $\mathbb{X}_a = f^{-1}(-\infty, a]$ to denote the sublevel sets of $\mathbb{X}$ defined by the value $a$. 

@@

 Given a  filtration $K$ with $m$ simplices and maximal dimension $d$, 
 one assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to the filtration, forming the \emph{filtration boundary matrix} $D$. 
 The reduction algorithm can  compute the persistent homology for all dimensions up to $d - 1$.
 In this case,  $D$ is a square matrix of size $m \times m$. Alternatively, a single dimension $i \leq d - 1$ can be computed, in which case one reduces a pair of sub-matrices  $D_i$ and $ D_{i+1}$ of $D$,  where $D_i$ has dimension $d_{i-1} \times d_i$ and $D_{i+1}$ is  $d_{i} \times d_{i+1}$. 
 If one is interested in a particular dimension of homology, then the latter approach is preferred. However, the  description of the algorithms is simpler when considering the full matrix, so $D$ will always be $(m \times m)$ in what follows. 
 All algorithms discussed generalize to both situations. We give an example using the matrix pair $(R_1, V_1)$  below. 

@@

Using $\mathbb{Z}_2$ coefficients and homology in dimension 1 to simplify the presentation, the reduction proceeds as follows:

@@

\noindent The column $c$ in $R$ indicates that dimension $1$ homology is born, which in this filtration is never killed since the dimension of the complex is $1$. Similarly, the columns at $u, v, w$ in $R_0$ (not shown) are all zero, indicating three $0$-dimensional homology classes  are born, which are then killed by the pivot entries in columns $a$ and $b$ in $R_1$.

@@

It is clear from Algorithm~\ref{alg:reduce} that a loose upper bound for this reduction is $O(m^3)$, where $m$ is the number of simplices of the filtration; it turns out that this bound is in fact tight, see~\cite{morozov2005persistence} for more details. 
There are actually many variations and optimizations proposed over the past decade to Algorithm~\ref{alg:reduce}; for example, one can   reverse the order of the filtration and compute cohomology~\cite{de2011dualities} instead. Despite these variations in computation, regardless of how the decomposition is obtained, it must obey the following invariants:

@@

 If these two invariants are satisfied, then we call the decomposition \emph{valid}. Given any valid $R = D V$ decomposition, then~\cite{zomorodian2005computing}:
%\begin{proposition}[Persistent Pairing~\cite{zomorodian2005computing}]
%	Let $R = D V$, where $R$ is in reduced form and $V$ is full-rank upper triangular (invariants (1) and (2)  hold). Then the pairing $(\sigma_i, \sigma_j)$ defined by $i = \mathrm{low}_R(j)$ is a \emph{creator/destroyer (persistent) pair}. While the matrices $R,V$ satisfying (1) and (2) are not unique in general,  the persistent pairs $(i,j)$ are unique and define the persistence diagrams of $K$.
%\end{proposition}
The collection of persistent pairings is independent of the choice of $R$ and $V$. When reducing the boundary matrix, we can therefore perform column operations in any order, as long as columns are added from left to right, and so long as every column operation performed in $R$ is also performed on $V$. 
Once $R$ is reduced, then the low entries in $R$ yield the correct pairings encoding the persistence diagram of the corresponding filtration $K$.

@@

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, and to detect homological critical events during the construction.
Notice that homological critical events can only occur when the filtration order changes, though the converse is not true.
Thus, detecting  all such events reduces to computing matrix decompositions at a finite set of time points interleaved between changes in the filtration order. 
At the finest scale, these changes manifest as transpositions of adjacent simplices. 
As a result, a fixed set of rules for maintaining a valid $R = D V$ decomposition under adjacent transpositions  is enough to simulate persistence dynamically. 
These rules prescribe certain column and row operations which must be applied to the matrix decomposition either before, during, or after the transposition, in order to ensure decomposition invariants I1 and I2 are respected.

Let $S_{i}^j$ represent the upper-triangular matrix such that multiplying by it on the right is equivalent to adding column $i$ to column $j$. 
Multiplication by $S_{i}^j$ on the left is equivalent to adding row $j$ to row $i$.
% NOT SURE about this:, and thus $S_{i}^j S_{i}^j = I$. 
Similarly, let $P$ denote the matrix so that multiplication from the right applies a fixed permutation to the columns of $A$.

@@

%We refer to the algorithm that maintains the $R = \partial V$ decomposition under adjacent permutations $P$ as the \emph{transposition framework}. 
% The explicit pseudocode describing these rules is given in the appendix by Algorithm~\ref{alg:tr}. 
%	\; = 
%	\begin{blockarray}{cccc}
%	\tiny{D_1} \kern-0.2em & a & b & c  \\
%  		u \kern-0.1em & \; 1 & 1 & 0 \; \topstrut \\
%  		v \kern-0.1em & \; 0 & 1 & 0 \; \\
%  		w \kern-0.1em & \; 1 & 0 & 0 \; \botstrut \\
%	\;\, 
%	+

>>>>

\section{Introduction} 
\subsection{Overview}\label{sec:overview} 
Given a triangulable topological space equipped with a tame continuous function, persistent homology captures the changes in topology across the sublevel sets of the space, and encodes them in a persistence diagram. The stability of persistence contends that   if the function changes continuously, so too will the points on the persistence diagram~\cite{cohen2007stability, cohen2006vines}. 
This motivates the  application of persistence to time-varying  settings, like that of dynamic metric spaces~\cite{kim2020spatiotemporal}. 
As persistence-related computations tend to exhibit high algorithmic complexity---essentially cubic\footnote{For finite fields, it is known that the persistence computation reduces to the PLU factorization problem, which takes $O(m^\omega)$ where $\omega \approx 2.373$ is the matrix multiplication constant.} in the size of the underlying filtration~\cite{morozov2005persistence}---their adoption to dynamic settings poses a challenging computational problem.
% With the current state of the art tools, there is no recourse to the problem of computing persistence from a time-varying filtration containing millions of simplices across thousands of snapshots in time.
Acquiring such a capability has far-reaching consequences: methods that vectorize persistence diagrams for machine learning purposes, like adaptive template functions~\cite{polanco2019adaptive} and persistence images~\cite{adams2017persistence}, or summary statistics like $\alpha$-smoothed Betti curves~\cite{ulmer2019topological}, all immediately become computationally viable tools in dynamic settings. 
%This analogy provides a strong interpretation at gaining insight on the subtle topological changes that may occur in studying a continuous process.

Cohen-Steiner et al. refer to a continuous 1-parameter family of persistence diagrams as a \emph{vineyard}, and they give in \cite{cohen2006vines} an efficient algorithm for their computation. 
The vineyards approach can be interpreted as an extension of the \emph{reduction} algorithm~\cite{zomorodian2005computing}, which computes the persistence diagram of a fixed filtration $K$ with $m$ simplices in $O(m^3)$ time via a decomposition $R = D V$ of the boundary matrix $D$ of $K$.
%Since $V$ is always full rank and upper-triangular, an alternative is to compute a $D = RU$ decomposition, where $U = V^{-1}$. 
In particular, the vineyards algorithm transforms a time-varying filtration into a particular set of permutations of the decomposition $R = DV$, each of which takes at most $O(m)$ time to execute. If one is interested in understanding how the persistent homology of a continuous function changes over time, then this algorithm is sufficient, for homological critical points can only occur when the filtration order changes. 
The vineyards algorithm is efficient asymptotically: if there are $d$ time-points where the filtration order changes, then  vineyards   takes $O(m^3 + md)$ time; one $O(m^3)$-time reduction upfront for the filtration at time $t_0$ followed by one $O(m)$ operation to update the decomposition for the remaining time points $(t_1, t_2, \dots, t_d)$. When $d >> m$, this $O(md)$ approach is far more efficient than the $O(dm^3)$  ``naive'' strategy of computing the diagrams independently at every time point.

@@

This work seeks to further understand and remedy this discrepancy: building on the work presented in~\cite{busaryev2010tracking}, we introduce a coarser approach to the vineyards algorithm.
While vineyards is  designed  for handling   \emph{continuous} time families of diagrams, it is not necessarily efficient when the time parameter is coarsely discretized.
Our methodology is based on the observation that practitioners often don't need (or want!) all of the persistence diagrams generated by a continuous 1-parameter of filtrations; usually just $n << d$ of them  suffice.   
% in many time varying settings, a valid decomposition is often only sought after at a relatively sparse number of time points relative to the size of the (finite) set needed to generated by the family. 
By exploiting the ``donor'' concept introduced in~\cite{busaryev2010tracking}, we are able to make a tradeoff between the number of times the decomposition is restored to a valid state and the granularity of the decomposition repair step, reducing the total number of column operations needed to apply an arbitrary permutation to the filtration. This trade off, paired with a fast greedy heuristic explained in section~\ref{sec:proxy_objective}, yields an algorithm that can update a $R = DV$ decomposition more efficiently than vineyards in coarse time-varying contexts, making dynamic persistence more computationally tractable for a wider class of use-cases. 
The source code containing both the algorithm we propose and the experiments performed in Section~\ref{sec:results} is open source and available online.\footnote{ Code is available here: \url{https://github.com/peekxc/dart}}

@@

  Consider the situation where one is concerned with capturing these dramatic changes in homology.
  On the left side of Figure~\ref{fig:vineyards}, we compare the cumulative cost (in total column operations) of various approaches.
  Since it is unknown a priori at when the persistent pairing function changes, one solution is to discretize the time domain at $n$ evenly spaced points and compute the diagrams at each time point independently. 
  An alternative approach is to construct a (e.g., straight-line) homotopy between  contiguous time-points, and then decompose these homotopies into adjacent transpositions in the filtration order---the vineyards approach. The former is often used in practice, but it is only an approximation; the latter is guaranteed to capture all homological changes in persistence that occur in simulating the 1-parameter family. 
  The graph in Figure~\ref{fig:vineyards} depicts two different size approximations (purple, $n=100$ and green, $n=10$), wherein the reduction algorithm (\texttt{pHcol}) was applied at $n$ time points, and two exact strategies (black and blue), which compute all $\approx 7 \times 10^{4}$ diagrams given by the homotopy. 
  As shown in the figure, if one needs to simulate the homotopy exactly, then the vineyards approach is indeed far more efficient than naively applying the reduction algorithm independently at all time points.
  However, when the discretization of the time domain is coarse enough, the naive approach actually performs less column operations than the vineyards strategy, while still capturing the main events.

@@

\subsection{Main results}
Given a filtered simplicial complex $K$ with filtration function $f : K \to [m]$, denote by $R = D V$ the decomposition of its corresponding boundary matrix $D$ such that $R$ is reduced and $V$ is upper-triangular (see section~\ref{sec:reduction} for details).
 If one has a pair of filtrations $(K, f)$,  $(K', f')$ of equal size $m = \lvert K \rvert = \lvert K' \rvert$ and $R = D V$ has been computed for $K$, then it may be advantageous to use the information stored in ($R$, $V$) to reduce the computation of $R' = D' V'$. 
Given a permutation $P$ such that $D' = P^T D P$, such an update scheme has the form: 
$$ (\ast P^T \ast R \ast P \ast) = (P^T DP)(\ast P^T \ast V \ast P \ast) $$
where $\ast$ is substituted with elementary column operations that repair the permuted decomposition. It is known how to linearly interpolate $f \mapsto f'$ using $d \sim O(m^2)$ updates to the decomposition, where each update is comprised of at most two column operations~\cite{cohen2006vines}. Since each column operation takes $O(m)$, the complexity performing this update is $O(m^3)$, which is efficient if all $d$ decompositions are needed. Otherwise, if only $(R', V')$ is needed (and none of the $d -1$ other decomposition), updating $R \mapsto R'$ using the approach from~\cite{cohen2006vines} matches the complexity of computing $R' = D' V'$ independently.

We now summarize the main results obtained in this effort (Theorem~\ref{prop:sparsity_result}): suppose one has a sequence of permutations $\mathcal{S} = \left( s_1, s_2, \dots, s_d \right)$ yielding a corresponding sequence of decompositions:
  \begin{equation}\label{eq:rv_seq}
   	R = R_0 = D_0 V_0 \overset{s_1}{\to} D_1 V_{1} \overset{s_2}{\to} \dots \overset{s_d}{\to} D_d V_{d} = R_d = R'
 \end{equation}
 where each permutation $s_i$ is a particular type of cyclic permutation (see section~\ref{sec:moves_setting}). If $i_\ast < j_\ast$, our first result is that this sequence can be computed in using $O(d\kappa)$ column operations, where: 
\begin{equation}
	\quad \kappa = \sum\limits_{k = 1}^d \, \lvert \mathbb{I}_{k}\rvert + \lvert \mathbb{J}_{k}\rvert
\end{equation}
where the quantities $\lvert \mathbb{I}_{k} \rvert$ and $\lvert \mathbb{J}_{k} \rvert$ depend on the sparsity of the $V_k$ and $R_k$ matrices, respectively, and $d \sim O(m)$ is a constant that depends on how similar $K$ and $K'$ are. The advantage of this result is that it depends explicitly on the sparsity pattern of the decomposition itself, which we take advantage of in Section~\ref{sec:schedule_cost}. Moreover, if the vineyards algorithm requires $C$ column operations to obtain~\eqref{eq:rv_seq}, we show (Proposition~\ref{prop:factor2}) that our approach obtains the same sequence using at most $L$ column operations, where: 
\begin{equation}
	\tfrac{1}{2} \, C \leq L \leq C
\end{equation}
Our second result turns towards the computation of $\mathcal{S}$ itself, and towards bounding its size $\lvert \mathcal{S} \rvert = d$. 
The complexity of determining $\mathcal{S}$ depends on the type of permutations considered; many edit distances exhibit tight quadratic growth rates and tend to require at least quadratic time (or worse) to compute~\cite{labarre2013lower}. 
We consider a special set of cyclic permutations whose sequences scale linearly in $m$ for all permutations in $S_m$: surprisingly, we find that the \emph{smallest} such sequence of these permutations which map $K \mapsto K'$ has size $d$ (Proposition~\ref{prop:schedule_size}), where: 
\begin{equation}
	d = m - \lvert \mathrm{LCS}(K, K') \rvert 
\end{equation}
Moreover, we also show that the information needed to construct any $\mathcal{S}$ with optimal size $\lvert \mathcal{S} \rvert = d$ can be computed in $O(m \log \log m)$ preprocessing time and $O(m)$ memory. Although $d$ clearly is $O(m)$ for pathological inputs, we show that $d \sim m - \sqrt{m}$ in expectation for random filtrations (Corollary~\ref{cor:expectation}) is not too pessimistic for truly random filtrations, and we give empirical results suggesting $d$ is tends to be much smaller for time-varying filtrations.

@@

%\begin{proposition}\label{prop:factor2_main}
%	Let $R = D V$ be a valid decomposition of size $(m \times m)$, let $C$ denote the number of $O(m)$ operations required to execute schedule $S_F$, and similarly let $L$ be the number of $O(m)$ operations required to execute a coarsened schedule $\widetilde{S}_F$ of $S_F$. We have the following relationship between $C$ and $L$:
%	$$ \frac{C}{2} \leq L \leq C$$ 
%\end{proposition}
\noindent 
\textbf{Outline:} The remainder of the paper is organized as follows: we review and establish the notations we will use to describe simplicial complexes, persistent homology, and dynamic persistence in Section \ref{sec:background}. 
We also cover the reduction algorithm (designated here as \texttt{pHcol}), the vineyards algorithm, and the set of \emph{move}-related algorithms introduced in~\cite{busaryev2010tracking}, which serves as the starting point of this work. 
In Section \ref{sec:move_schedules} we introduce our move schedules  and provide efficient algorithms to make this alternative viable. 
In Section \ref{sec:results} we present applications of the proposed method. 
In particular, to the computation of crocker stacks from flock simulations, and to 2-dimensional persistence on the Mumford data set of image patches from natural images. 
In Section \ref{sec:conclusion} we conclude the paper by discussing other possible applications and future work.

@@

Suppose one has a family $\{K_i\}_{i\in I}$ of simplicial complexes indexed by a totally ordered   set $I$, and so that for any $i< j \in I$ we have $K_i \subseteq K_j$. There are two index sets of interest in what follows: $\mathbb{R}$ and $[n] = \{ 1, \dots, n\}$. 
Such a family is called a \emph{filtration},
which is deemed \emph{simplexwise} if $K_j \smallsetminus K_i = \{\sigma_j\}$ 
whenever $j$ is the immediate successor of $i$ in $I$.
Any finite filtration may be trivially converted into an simplexwise filtration via a set of \emph{condensing}, \emph{refining}, and \emph{reindexing} maps (see~\cite{bauer2021ripser} for more details). 
Without loss of generality, we exclusively consider simplexwise filtrations but for brevity-sake refer to them simply as filtrations.

@@

\noindent \textbf{Example 2.1:} Non-simplexwise filtrations often arise in geometrical contexts. For example, given a finite metric space $(X, d)$, the \emph{Vietoris-Rips} complex at scale $\epsilon \in \mathbb{R}$ is the abstract simplicial complex given by: 

@@

A VR filtration indexed over $\mathbb{R}$ can be condensed to an simplexwise filtration indexed over the set of pairwise distances $\{\, d(x,x') \; \mid \; x,x' \in X \, \}$  by extending a total order of $X$ to its power set using any linear extension of the face poset, such as e.g. the lexicographical ordering. 

@@

Let $\mathbb{X}$ be a triangulable topological space and $K$ an abstract simplicial complex whose geometric realization is homeomorphic to $\mathbb{X}$. 
Let  $f: \mathbb{X} \to \mathbb{R}$ be continuous and  write $\mathbb{X}_a = f^{-1}(-\infty, a]$ to denote the sublevel sets of $\mathbb{X}$ defined by the value $a$. 

@@

 Given a filtration $K$ with $m$ simplices, one assembles the elementary boundary chains $\partial(\sigma)$ as columns ordered according to the filtration, forming the square $m \times m$  \emph{filtration boundary matrix} $D$. 
 The reduction algorithm computes the persistent homology from $D$ for all dimensions up to $\mathrm{dim}(K) - 1$.
  Alternatively, a single dimension $p \leq \mathrm{dim}(K) - 1$ can be computed, in which case one reduces a pair of sub-matrices $D_p$ and $ D_{p+1}$ of $D$,  where $D_p$ has dimension $d_{p-1} \times d_p$ and $D_{p+1}$ is $d_{p} \times d_{p+1}$. 
In some situations the description of the ensuing algorithms is simpler when considering all dimensions simultaneously, while in others it is more succinct to consider dimension-specific matrices. 
In what follows $D$, $R$, and $V$ will always denote square $(m \times m)$ matrices, though we note that all algorithms discussed generalize to both cases. 
 We give an example of the reduction algorithm using the matrix pair $(R_1, V_1)$ below. 

@@

Using $\mathbb{Z}_2$ coefficients and homology in dimension 1 for simplicity, the reduction proceeds as follows:

@@

\noindent Since column $c$ in $R_1$ is $0$, the chain $d_1 + d_2 + d_3$ indicated by the column $c$ in $V_1$ represents a dimension $1$ cycle. Similarly, the columns at $u, v, w$ in $R_0$ (not shown) are all zero, indicating three $0$-dimensional homology classes are born, two of which are killed by the pivot entries in columns $a$ and $b$ in $R_1$.

@@

Inspection of Algorithm~\ref{alg:reduce} suggests that a loose upper bound for the reduction is $O(m^3)$, where $m$ is the number of simplices of the filtration---this bound is in fact tight~\cite{morozov2005persistence}. 
Nonetheless, many variations and optimizations to Algorithm~\ref{alg:reduce} have been proposed over the past decade, see~\cite{bauer2017phat, bauer2021ripser, chen2011persistent} for an overview. Despite these variations in computation, the resulting decomposition must satisfy the following invariants:

@@

 If these two invariants are satisfied, then we call the decomposition \emph{valid}. 
%%\begin{proposition}[Persistent Pairing~\cite{zomorodian2005computing}]
%%	Let $R = D V$, where $R$ is in reduced form and $V$ is full-rank upper triangular (invariants (1) and (2)  hold). Then the pairing $(\sigma_i, \sigma_j)$ defined by $i = \mathrm{low}_R(j)$ is a \emph{creator/destroyer (persistent) pair}. While the matrices $R,V$ satisfying (1) and (2) are not unique in general,  the persistent pairs $(i,j)$ are unique and define the persistence diagrams of $K$.
%%\end{proposition}
Though the matrices $R$ and $V$ are not unique, it has been shown that the collection of persistent pairings (derived from $R$) are~\cite{zomorodian2005computing}. 
%When reducing the boundary matrix, we can therefore perform column operations in any order, as long as columns are added from left to right, and so long as every column operation performed in $R$ is also performed on $V$. 
In particular, once $R$ is in reduced form, the persistence diagram of the corresponding filtration $K$ is given by lowest non-zero ``pivot'' entries in $R$.

@@

The original purpose of the vineyards algorithm, as described in~\cite{cohen2006vines}, was to compute a continuous 1-parameter family of persistence diagrams over a time-varying filtration, detecting homological critical events along the way.
As homological critical events can only occur when the filtration order changes, detecting all such critical events can be reduced to computing valid decompositions at all time points interleaving  changes in the filtration order. 
At the finest scale, these changes manifest as transpositions of adjacent simplices for simplexwise filtrations. 
As a result, any fixed set of rules that maintains a valid $R = D V$ decomposition under adjacent transpositions is sufficient to simulate persistence dynamically. 
These rules prescribe certain column and row operations to apply to a given matrix decomposition either before, during, or after each individual transposition to ensure it is a valid decomposition.

Let $S_{i}^j$ represent the upper-triangular matrix such that $A S_{i}^j$ is equivalent to the elementary operation that adds column $i$ of $A$ to column $j$ of $A$ and $S_{i}^j A$ is equivalent to adding row $j$ of $A$ to row $i$ of $A$.
Similarly, let $P$ denote the matrix so that $A P$ applies a fixed permutation to the columns of $A$ and $P^T A$ applied the same permutation to the rows of $A$.

@@
